### 2015-04-07 contract-language.tex
cf to rcf. Better formatting for operation names.

The syntax of our core contract language is shown in Figure
~\ref{fig:contract-lang}. The language is based on first-order logic (FOL), and
admits prenex universal quantification over typed and untyped effect variables.
We use a special effect variable ($\cureff$) to denote the effect of
\emph{current operation} - the operation for which a contract is being written.
Notice that $\cureff$ occurs free in the contract. We will fix its scope when
classifying the contracts (\S~\ref{sec:classify}). The type of an effect is
simply the name of the operation (eg: \cf{withdraw}) that induced the effect.
We admit disjunction in types to let an effect variable range over multiple
operation names. The contract $\small \forall (a : \tau_1 \vee \tau_2).~\psi$
is just syntactic sugar for $\small \forall a. (\oper{a}{\tau_1} \vee
\oper{a}{\tau_2}) \Rightarrow \psi$. An untyped effect variable ranges over all
operation names.

@@

\item \textbf{Eventual consistency}: Eventually consistent operations can
	be satisfied as long as the client can reach at least one replica. In the
	bank account example, \cf{deposit} is an eventually consistent operation.
	While an \ecds typically offer \emph{basic} eventual consistency with all
	possible anomalies, we assume that our store provides stronger semantics that
	remain highly-available~\cite{BailisHAT,COPS}; the store always exposes a
	\emph{causal cut} of the updates. This semantics can be formally captured in
	terms of the following contract definition:

@@

\item \textbf{Causal consistency}: Causally consistent operations are required
	to see a causally consistent snapshot of the object state, including the
	actions performed on the same session.  The latter requirement implies that
	if two operations $o_1$ and $o_2$ from the same session are applied to two
	different replicas $r_1$ and $r_2$, the second operation cannot be discharged
	until the effect of $o_1$ is included in $r_2$. The \cf{getBalance} operation
	requires causal consistency, as it requires the operations from the same
	session to be visible, which cannot be guaranteed under eventual consistency.
	The corresponding store semantics is captured by the contract $\ccc$ defined
	below:

@@

\item \textbf{Strong consistency}: Strongly consistent operations may block
  indefinitely under network partitions. An example is the total-order
  contract on \cf{withdraw} operation. The corresponding store semantics is
	captured by the $\scc$ contract definition:

@@

The classification scheme, presented formally in Figure~\ref{sem:classify},
defines rules to judge the consistency class of a contact. For example, the
scheme classifies the \cf{getBalance} contract ($\cv_{gb}$) from
\S~\ref{sec:motivation} as a {\sf\small CausallyConsistent} contract, because
the sequent $\Delta \vdash \ccc \Rightarrow \cv_{gb}$ is valid in first-order
logic (therefore, $\cv_{gb} \le \ccc$), whereas the sequent $\Delta \vdash \ecc
\Rightarrow \cv_{gb}$ is invalid (therefore, $\cv_{gb} \not\le \ecc$). Since we
confine our contract language to a decidable subset of the logic, validity
of such sequents can be decided mechanically allowing us to automate the
classification scheme in \name.

@@

Here, the consistency level for an operation in this scheme is any combination
of the above guarantees. It is easy to see that there are 24 (4!) combinations,
which form a partially ordered consistency lattice with a height of 4.
Classifying a contract under this scheme amounts to determining the weakest
combination under which the contract will hold, and entails 24 comparisons at
worst. Under this scheme, \cf{deposit} operations does not need any guarantees,
\cf{getBalance} needs RYW and WFR ($\Delta \vdash \R{RYW} \wedge \R{WFR}
\Rightarrow \cv_{gb}$), and \cf{withdraw} cannot be satisfied ($\Delta \vdash
\R{RYW} \wedge \R{MW} \wedge \R{MR} \wedge \R{WFR} \not\Rightarrow \cv_{w}$).

>>>>

The syntax of our core contract language is shown in Figure
~\ref{fig:contract-lang}. The language is based on first-order logic (FOL), and
admits prenex universal quantification over typed and untyped effect variables.
We use a special effect variable ($\cureff$) to denote the effect of
\emph{current operation} - the operation for which a contract is being written.
Notice that $\cureff$ occurs free in the contract. We will fix its scope when
classifying the contracts (\S~\ref{sec:classify}). The type of an effect is
simply the name of the operation (eg: \rcf{withdraw}) that induced the effect.
We admit disjunction in types to let an effect variable range over multiple
operation names. The contract $\small \forall (a : \tau_1 \vee \tau_2).~\psi$
is just syntactic sugar for $\small \forall a. (\oper{a}{\tau_1} \vee
\oper{a}{\tau_2}) \Rightarrow \psi$. An untyped effect variable ranges over all
operation names.

@@

\item \textbf{Eventual consistency}: Eventually consistent operations can
	be satisfied as long as the client can reach at least one replica. In the
	bank account example, \rcf{deposit} is an eventually consistent operation.
	While an \ecds typically offer \emph{basic} eventual consistency with all
	possible anomalies, we assume that our store provides stronger semantics that
	remain highly-available~\cite{BailisHAT,COPS}; the store always exposes a
	\emph{causal cut} of the updates. This semantics can be formally captured in
	terms of the following contract definition:

@@

\item \textbf{Causal consistency}: Causally consistent operations are required
	to see a causally consistent snapshot of the object state, including the
	actions performed on the same session.  The latter requirement implies that
	if two operations $o_1$ and $o_2$ from the same session are applied to two
	different replicas $r_1$ and $r_2$, the second operation cannot be discharged
	until the effect of $o_1$ is included in $r_2$. The \rcf{getBalance} operation
	requires causal consistency, as it requires the operations from the same
	session to be visible, which cannot be guaranteed under eventual consistency.
	The corresponding store semantics is captured by the contract $\ccc$ defined
	below:

@@

\item \textbf{Strong consistency}: Strongly consistent operations may block
  indefinitely under network partitions. An example is the total-order
  contract on \rcf{withdraw} operation. The corresponding store semantics is
	captured by the $\scc$ contract definition:

@@

The classification scheme, presented formally in Figure~\ref{sem:classify},
defines rules to judge the consistency class of a contact. For example, the
scheme classifies the \rcf{getBalance} contract ($\cv_{gb}$) from
\S~\ref{sec:motivation} as a {\sf\small CausallyConsistent} contract, because
the sequent $\Delta \vdash \ccc \Rightarrow \cv_{gb}$ is valid in first-order
logic (therefore, $\cv_{gb} \le \ccc$), whereas the sequent $\Delta \vdash \ecc
\Rightarrow \cv_{gb}$ is invalid (therefore, $\cv_{gb} \not\le \ecc$). Since we
confine our contract language to a decidable subset of the logic, validity
of such sequents can be decided mechanically allowing us to automate the
classification scheme in \name.

@@

Here, the consistency level for an operation in this scheme is any combination
of the above guarantees. It is easy to see that there are 24 (4!) combinations,
which form a partially ordered consistency lattice with a height of 4.
Classifying a contract under this scheme amounts to determining the weakest
combination under which the contract will hold, and entails 24 comparisons at
worst. Under this scheme, \rcf{deposit} operations does not need any guarantees,
\rcf{getBalance} needs RYW and WFR ($\Delta \vdash \R{RYW} \wedge \R{WFR}
\Rightarrow \cv_{gb}$), and \rcf{withdraw} cannot be satisfied ($\Delta \vdash
\R{RYW} \wedge \R{MW} \wedge \R{MR} \wedge \R{WFR} \not\Rightarrow \cv_{w}$).

### 2015-04-07 evaluation.tex
Use correct class file for the camera ready version. Remove vertical space adjustments.
Formatting changes for camera-ready
Formatting edits
cf to rcf. Better formatting for operation names.

\begin{itemize}[noitemsep]
\item \textbf{LWW register}: A last-write-wins register that provides read
	and write operations, where the read returns the value of the latest write.

@@

\begin{table}
\setlength{\tabcolsep}{4pt}
{\sffamily \small
\begin{center}
\begin{tabular} {|l|r|r|r|r|r|r|r|r|}
\hline
{\bf Benchmark} & {\bf LOC} & {\bf \#T} & {\bf EC} & {\bf CC} & {\bf SC} & {\bf RC} & {\bf MAV} & {\bf RR} \\
\hline
{LWW Reg} & 108 & 1 & 2 & 2 & 2 & 0 & 0 & 0 \\
{DynamoDB} & 126 & 1 & 3 & 1 & 2 & 0 & 0 & 0 \\
{Bank Account} & 155 & 1 & 1 & 1 & 1 & 1 & 0 & 1 \\
{Shopping List} & 140 & 1 & 2 & 1 & 1 & 0 & 0 & 0 \\
{Online store} & 340 & 4 & 9 & 1 & 0 & 2 & 0 & 1 \\
{RUBiS} & 640 & 6 & 14 & 2 & 1 & 4 & 2 & 0 \\
{Microblog} & 659 & 5 & 13 & 6 & 1 & 6 & 3 & 1 \\
\hline
\end{tabular}
\end{center} }
\caption{The distribution of classified contracts. \#T refers to the number of
tables in the application. The columns 4-6 (7-9) represent operations
(transactions) assigned to this consistency (isolation) level.}
\label{tab:ctrts}
\end{table}

@@

\begin{figure*}
  \vspace{-5pt}
  \centering
  \subfigure[Bank account]{\label{grf:BA-tp-vs-lat}\includegraphics[width=0.24\textwidth]{graphs/BA-tp-vs-lat.pdf}}
  \subfigure[LWW transactions]{\label{grf:LWW-txn}\includegraphics[width=0.24\textwidth]{graphs/LWW-txn.pdf}}
  \subfigure[RUBiS bidding mix]{\label{grf:rubis}\includegraphics[width=0.24\textwidth]{graphs/Rubis.pdf}}
  \subfigure[Impact of summarization]{\label{grf:summarization}\includegraphics[width=0.235\textwidth]{graphs/summarization.pdf}}
	\caption{Quelea Performance.}
  \label{grf:LWW_perf}
\end{figure*}

@@

The lines marked EC and CC correspond to all operations (including
\cf{withdraw}) being assigned EC and CC consistency levels. These levels
compromise correctness as \cf{withdraw} has to be an SC operation. The SC line
corresponds to a configuration where all operations are strongly consistent;
this ensures application correctness, at the cost of performance. \name
corresponds to our implementation, which classifies operations based on their
contract specifications. With 512 clients, the \name implementation was within
41\% of the latency and 18\% of the throughput of EC, whereas SC operations had
162\% higher latency and 52\% lower throughput than EC operations. Observe that
there is a point in each case after which the latency increases while the
throughput decreases; these correspond to points where the store becomes
saturated with client requests. In a \cf{2DC} configuration (not shown here),
the average latency of SC operations with 512 clients increased by 9.4$\times$
due to the cost of geo-distributed coordination, whereas \name operations were
only 2.2$\times$ slower, mainly due to the increased cost of \cf{withdraw}
operations. Importantly, the latency of \cf{getBalance} and \cf{deposit}
remained almost the same, illustrating the benefit of fine-grained contract
classification.

>>>>

\begin{itemize}
\item \textbf{LWW register}: A last-write-wins register that provides read
	and write operations, where the read returns the value of the latest write.

@@

\begin{table}
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.2}
{\sffamily
\begin{center}
\caption{The distribution of classified contracts. \#T refers to the number of
tables in the application. The columns 4-6 (7-9) represent operations
(transactions) assigned to this consistency (isolation) level.}
\label{tab:ctrts}
\begin{tabular} {rrrrrrrrr}
\hline
{\bf Benchmark} & {\bf LOC} & {\bf \#T} & {\bf EC} & {\bf CC} & {\bf SC} & {\bf RC} & {\bf MAV} & {\bf RR} \\
\hline
{LWW Reg} & 108 & 1 & 2 & 2 & 2 & 0 & 0 & 0 \\
{DynamoDB} & 126 & 1 & 3 & 1 & 2 & 0 & 0 & 0 \\
{Bank Account} & 155 & 1 & 1 & 1 & 1 & 1 & 0 & 1 \\
{Shopping List} & 140 & 1 & 2 & 1 & 1 & 0 & 0 & 0 \\
{Online store} & 340 & 4 & 9 & 1 & 0 & 2 & 0 & 1 \\
{RUBiS} & 640 & 6 & 14 & 2 & 1 & 4 & 2 & 0 \\
{Microblog} & 659 & 5 & 13 & 6 & 1 & 6 & 3 & 1 \\
\hline
\end{tabular}
\end{center} }
\end{table}

@@

\begin{figure*}
  \centering
  \subfigure[Bank account]{\label{grf:BA-tp-vs-lat}\includegraphics[width=0.24\textwidth]{graphs/BA-tp-vs-lat.pdf}}
  \subfigure[LWW transactions]{\label{grf:LWW-txn}\includegraphics[width=0.24\textwidth]{graphs/LWW-txn.pdf}}
  \subfigure[RUBiS bidding mix]{\label{grf:rubis}\includegraphics[width=0.24\textwidth]{graphs/Rubis.pdf}}
  \subfigure[Impact of summarization]{\label{grf:summarization}\includegraphics[width=0.235\textwidth]{graphs/summarization.pdf}}
	\caption{Quelea Performance.}
  \label{grf:LWW_perf}
\end{figure*}

@@

The lines marked EC and CC correspond to all operations (including
\rcf{withdraw}) being assigned EC and CC consistency levels. These levels
compromise correctness as \rcf{withdraw} has to be an SC operation. The SC line
corresponds to a configuration where all operations are strongly consistent;
this ensures application correctness, at the cost of performance. \name
corresponds to our implementation, which classifies operations based on their
contract specifications. With 512 clients, the \name implementation was within
41\% of the latency and 18\% of the throughput of EC, whereas SC operations had
162\% higher latency and 52\% lower throughput than EC operations. Observe that
there is a point in each case after which the latency increases while the
throughput decreases; these correspond to points where the store becomes
saturated with client requests. In a \cf{2DC} configuration (not shown here),
the average latency of SC operations with 512 clients increased by 9.4$\times$
due to the cost of geo-distributed coordination, whereas \name operations were
only 2.2$\times$ slower, mainly due to the increased cost of \rcf{withdraw}
operations. Importantly, the latency of \rcf{getBalance} and \rcf{deposit}
remained almost the same, illustrating the benefit of fine-grained contract
classification.

### 2015-04-07 motivation.tex
Use correct class file for the camera ready version. Remove vertical space adjustments.
Formatting changes for camera-ready
Formatting edits
cf to rcf. Better formatting for operation names.

\subsection{RDT specification}

@@

Let us assume that the bank account object provides three operations:
\cf{deposit}, \cf{withdraw} and \cf{getBalance}, with the assumption that the
withdraw fails if the account has insufficient balance. Every operation in
\name is of the following type, written in Haskell syntax:

@@

\begin{figure}
\vspace{-2.3mm}
\begin{codehaskell}
data Acc = Deposit Int | Withdraw Int | GetBal

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw hist v =
	if sel1 dollar getBalance hist () >= v
  then (True, Just dollar Withdraw v)
	else (False, Nothing)
\end{codehaskell}
\caption{Definition of a bank account expressed in Quelea.}
\label{fig:ex}
\end{figure}

The datatype \cf{Acc} represents the effect type for the bank account. The
function \cf{sum} returns the sum of elements in the list, and \cf{sel1}
returns the first element of a tuple. For each operation, \cf{hist} is a
\emph{snapshot} of the state of the object at some replica. In this sense,
every operation on the RDT is atomic, and thus amenable to sequential reasoning.
Here, \cf{getBalance} is a read-only operation,
\cf{deposit} always emits an effect, and \cf{withdraw} only emits an effect if
there is sufficient balance in the account. We have implemented a large corpus
of RDTs for realistic benchmarks including shopping carts, auction and
micro-blogging sites, etc. in a few tens of lines of code, expressed in this style.

@@

Observe that the definition of \cf{getBalance} reduces over the \emph{entire
history} of updates to an account. If we are to realize an efficient
implementation of this bank account RDT, we need a \emph{summary} of the
account history. Intuitively, the current account balance summarizes the state
of an account. A bank account with the history \cf{[Deposit 10, Withdraw 5]} is
\emph{observably equivalent} to a bank account with a single deposit operation
\cf{[Deposit 5]}; we can replace the earlier history with the latter and a
client of the store would not able to tell the difference between the two.

@@

{\small
\begin{verbatim}
 summarize hist = 
   [Deposit $ sel1 $ getBalance hist ()]
\end{verbatim}}

@@

Our goal is to choose the correct consistency level for each of the bank
account operations such that (1) the balance remains non-negative and (2) the
\cf{getBalance} operation never incorrectly returns a negative balance.

\begin{figure}[ht]
\vspace{-2mm}
\centering
\subfigure[Unsafe withdraw]{\label{fig:unsafeWithdrawAnomaly}\includegraphics[width=0.34\columnwidth]{Figures/Motivation4}}
\hfill
\subfigure[Negative balance]{\label{fig:negativeBalanceAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation2}}
\hfill
\subfigure[Missing update]{\label{fig:missingUpdateAnomaly}\includegraphics[width=0.26\columnwidth]{Figures/Motivation1}}
\caption{Anomalies possible under eventual consistency for the get balance operation.}
\label{fig:ba_anomalies}
\end{figure}

Consider the execution shown in Figure~\ref{fig:unsafeWithdrawAnomaly}. Assume
that all operations in the figure are on the same bank account object with the
initial balance being zero. Session 1 performs a \cf{deposit} of 100, followed
by a \cf{withdraw} of 80 in the same session. The \cf{withdraw} operation
witnesses the deposit and succeeds\footnote{Although visibility and session
order relations relate effects, we have abused the notation in these examples
to relate operations, with the idea that the relations relate the effect
emitted by those operations}. Subsequently, session 2 perform a \cf{withdraw}
operation, but importantly, due to eventual consistency, only witnesses the
\cf{deposit} from session 1, but not the subsequent withdraw. Hence, this
\cf{withdraw} also \emph{incorrectly} succeeds, violating the integrity
constraint. A subsequent \cf{getBalance} operation, that happens to witness all
the previous operations, would report a negative balance.

It is easy to see that preventing concurrent \cf{withdraw} operations
eliminates this anomaly. This can be done by insisting that \cf{withdraw} be
executed as a strongly consistent operation. Despite this strengthening,
the \cf{getBalance} operation may still incorrectly report a negative balance to
the user. Consider the execution shown in
fig.~\ref{fig:negativeBalanceAnomaly}, which consists of three concurrent
sessions performing a \cf{deposit}, a \cf{withdraw}, and a \cf{getBalance}
operation, respectively, on the same bank account object. As the {\sf vis} edge
indicates, operation \cf{withdraw(50)} in session 2 witnesses the effects of
\cf{deposit(100)} from session 1, concludes that there is sufficient balance,
and completes successfully. However, the \cf{getBalance} operation may only
witness this successful withdraw, but not the causally preceding \cf{deposit},
and reports the balance of negative 50 to the user.

Under eventual consistency, the users may also be exposed to other forms of
inconsistencies. Figure~\ref{fig:missingUpdateAnomaly} shows an execution where
the \cf{getBalance} operation in a session does not witness the effects of an
earlier \cf{withdraw} operation performed in the same session, possibly because
it was served by a replica that has not yet merged the \cf{withdraw} effect.
This anomaly leads the user to incorrectly conclude that the \cf{withdraw}
operation failed to go through.

Although it is easy to understand the reasons behind the occurrence of the
aforementioned anomalies, finding the appropriate fixes is not readily
apparent. Making \cf{getBalance} a strongly consistent operation is definitely
sufficient to avert anomalies, but is it really necessary? Given the cost of
enforcing strong consistency~\cite{DynamoDB, Pileus}, it is preferable to avoid
it unless there are no viable alternatives. Exploring the space of these
alternatives requires understanding the subtle differences in semantics of
various kinds of weak consistency alternatives.

@@

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice-versa. We express this application-level consistency requirement as a
contract ($\cv_w$) over \cf{withdraw}:

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).~\sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{smathpar}

\noindent Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect  emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

\noindent For \cf{getBalance}, we construct the following contract ($\cv_{gb}$):

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad (\vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}) \\
\qquad \wedge~ ((\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff})
\end{array}
\end{smathpar}

\noindent The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$
is an effect that was emitted either by a \cf{deposit} or a \cf{withdraw}
operation. If a \cf{withdraw} $b$ is visible to \cf{getBalance} $\cureff$, then
all \cf{deposit} operations $a$ visible to $b$ should also be visible to
$\cureff$. Intuitively, this prevents negative balance anomalies. Our contract
language provides operators to compose relations. The syntax $(R_1 \cap
R_2)(a,b)$ is equivalent to $R_1(a,b) \wedge R_2(a,b)$. The last line of the
above contract says that if a \cf{deposit} or a \cf{withdraw} operation
precedes a \cf{getBalance} operation in session order, and is applied on the
same object as the \cf{getBalance} operation, then it must be the case that the
\cf{getBalance} operation witnesses the effects of the preceding operations.

Finally, since there are no restrictions on when or how a \cf{deposit}
operation can execute, its contract is simply $\small \true$.

@@

Notice that the contracts for \cf{withdraw} and \cf{getBalance} only express
application-level consistency requirements, and make no reference to the
semantics of the underlying store. To write contracts, a programmer only needs
to reason about the semantics of the application under the \name system model.
The mapping of application-level consistency requirements to appropriate
store-level guarantees is done automatically behind-the-scene. How might one go
about ensuring that an execution adheres to a contract? The challenge is that a
contract provides a declarative (axiomatic) specification of an execution,
while what is required is an operational procedure for \emph{enforcing} its
implicit constraints.

>>>>

\subsection{RDT Specification}

@@

Let us assume that the bank account object provides three operations:
\rcf{deposit}, \rcf{withdraw} and \rcf{getBalance}, with the assumption that the
withdraw fails if the account has insufficient balance. Every operation in
\name is of the following type, written in Haskell syntax:

@@

\begin{figure}
\begin{codehaskell}
data Acc = Deposit Int | Withdraw Int | GetBal

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw hist v =
	if sel1 dollar getBalance hist () >= v
  then (True, Just dollar Withdraw v)
	else (False, Nothing)
\end{codehaskell}
%\captionsetup{singlelinecheck=off}
\caption{Definition of a bank account expressed in Quelea.}
\label{fig:ex}
\end{figure}

The datatype \cf{Acc} represents the effect type for the bank account. The
function \cf{sum} returns the sum of elements in the list, and \cf{sel1}
returns the first element of a tuple. For each operation, \cf{hist} is a
\emph{snapshot} of the state of the object at some replica. In this sense,
every operation on the RDT is atomic, and thus amenable to sequential reasoning.
Here, \rcf{getBalance} is a read-only operation,
\rcf{deposit} always emits an effect, and \rcf{withdraw} only emits an effect if
there is sufficient balance in the account. We have implemented a large corpus
of RDTs for realistic benchmarks including shopping carts, auction and
micro-blogging sites, etc. in a few tens of lines of code, expressed in this style.

@@

Observe that the definition of \rcf{getBalance} reduces over the \emph{entire
history} of updates to an account. If we are to realize an efficient
implementation of this bank account RDT, we need a \emph{summary} of the
account history. Intuitively, the current account balance summarizes the state
of an account. A bank account with the history \cf{[Deposit 10, Withdraw 5]} is
\emph{observably equivalent} to a bank account with a single deposit operation
\cf{[Deposit 5]}; we can replace the earlier history with the latter and a
client of the store would not able to tell the difference between the two.

@@

\begin{codehaskell}
 summarize hist =
   [Deposit dollar sel1 dollar getBalance hist ()]
\end{codehaskell}

@@

Our goal is to choose the correct consistency level for each of the bank
account operations such that (1) the balance remains non-negative and (2) the
\rcf{getBalance} operation never incorrectly returns a negative balance.

\begin{figure}[ht]
\centering
\subfigure[Unsafe withdraw]{\label{fig:unsafeWithdrawAnomaly}\includegraphics[width=0.34\columnwidth]{Figures/Motivation4}}
\hfill
\subfigure[Negative balance]{\label{fig:negativeBalanceAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation2}}
\hfill
\subfigure[Missing update]{\label{fig:missingUpdateAnomaly}\includegraphics[width=0.26\columnwidth]{Figures/Motivation1}}
\caption{Anomalies possible under eventual consistency for the get balance operation.}
\label{fig:ba_anomalies}
\end{figure}

Consider the execution shown in Figure~\ref{fig:unsafeWithdrawAnomaly}. Assume
that all operations in the figure are on the same bank account object with the
initial balance being zero. Session 1 performs a \rcf{deposit} of 100, followed
by a \rcf{withdraw} of 80 in the same session. The \rcf{withdraw} operation
witnesses the deposit and succeeds\footnote{Although visibility and session
order relations relate effects, we have abused the notation in these examples
to relate operations, with the idea that the relations relate the effect
emitted by those operations}. Subsequently, session 2 perform a \rcf{withdraw}
operation, but importantly, due to eventual consistency, only witnesses the
\rcf{deposit} from session 1, but not the subsequent withdraw. Hence, this
\rcf{withdraw} also \emph{incorrectly} succeeds, violating the integrity
constraint. A subsequent \rcf{getBalance} operation, that happens to witness all
the previous operations, would report a negative balance.

It is easy to see that preventing concurrent \rcf{withdraw} operations
eliminates this anomaly. This can be done by insisting that \rcf{withdraw} be
executed as a strongly consistent operation. Despite this strengthening,
the \rcf{getBalance} operation may still incorrectly report a negative balance to
the user. Consider the execution shown in
fig.~\ref{fig:negativeBalanceAnomaly}, which consists of three concurrent
sessions performing a \rcf{deposit}, a \rcf{withdraw}, and a \rcf{getBalance}
operation, respectively, on the same bank account object. As the {\sf vis} edge
indicates, operation \rcf{withdraw(50)} in session 2 witnesses the effects of
\rcf{deposit(100)} from session 1, concludes that there is sufficient balance,
and completes successfully. However, the \rcf{getBalance} operation may only
witness this successful withdraw, but not the causally preceding \rcf{deposit},
and reports the balance of negative 50 to the user.

Under eventual consistency, the users may also be exposed to other forms of
inconsistencies. Figure~\ref{fig:missingUpdateAnomaly} shows an execution where
the \rcf{getBalance} operation in a session does not witness the effects of an
earlier \rcf{withdraw} operation performed in the same session, possibly because
it was served by a replica that has not yet merged the \rcf{withdraw} effect.
This anomaly leads the user to incorrectly conclude that the \rcf{withdraw}
operation failed to go through.

Although it is easy to understand the reasons behind the occurrence of the
aforementioned anomalies, finding the appropriate fixes is not readily
apparent. Making \rcf{getBalance} a strongly consistent operation is definitely
sufficient to avert anomalies, but is it really necessary? Given the cost of
enforcing strong consistency~\cite{DynamoDB, Pileus}, it is preferable to avoid
it unless there are no viable alternatives. Exploring the space of these
alternatives requires understanding the subtle differences in semantics of
various kinds of weak consistency alternatives.

@@

In our running example, it is clear that in order to preserve the integrity
constraint, the \rcf{withdraw} operation must be strongly consistent.  That is,
given two \rcf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice-versa. We express this application-level consistency requirement as a
contract ($\cv_w$) over \rcf{withdraw}:

\begin{cmathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).\\
\qquad \sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{cmathpar}

\noindent Here, $\cureff$ stands for the effect emitted by the \rcf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect  emitted
by a \rcf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \rcf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\rcf{withdraw} operation is said to be derived from a correct implementation of
\rcf{withdraw}.

To prevent \rcf{getBalance} from ever showing a negative balance, it is
necessary to prevent the scenario depicted in
fig.~\ref{fig:negativeBalanceAnomaly}. Let $\cureff$ stand for the effect
emitted by the \rcf{getBalance} operation. If the effect ($b$) of a withdraw
operation is visible to $\cureff$, and the effect ($a$) of a deposit operation
is visible to the effect ($b$) of the withdraw operation, then it must be the
case that $a$ is also visible to $\cureff$. A contract ($\cv_{gb}^1$) for
\rcf{getBalance} operation that precisely captures this application-level consistency
requirement can be written thus:

\begin{cmathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}). \\
\qquad (\vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff})
\end{array}
\end{cmathpar}

\noindent To prevent the missing update anomaly described in
fig.~\ref{fig:missingUpdateAnomaly}, it is necessary for a \rcf{getBalance}
operation on a bank account to witness the effects of all previous \rcf{deposit}
and \rcf{withdraw} operations performed on the same bank account in the same
session. We can express an additional contract ($\cv_{gb}^2$) for
\rcf{getBalance} that captures this consistency requirement:

\begin{cmathpar}
\begin{array}{l}
\forall (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad((\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff})
\end{array}
\end{cmathpar}

\noindent Our contract language provides operators to compose relations. The
syntax $(R_1 \cap R_2)(a,b)$ is equivalent to $R_1(a,b) \wedge R_2(a,b)$. The
above contract ($\cv_{gb}^2$) says that if a \rcf{deposit} or a \rcf{withdraw}
operation precedes a \rcf{getBalance} operation in session order, and is applied
on the same object as the \rcf{getBalance} operation, then it must be the case
that the \rcf{getBalance} operation witnesses the effects of the preceding
operations.

The final contract ($\cv_{gb}$) of the \rcf{getBalance} operation is
merely a conjunction of the previous two versions ($\cv_{gb}^1$ and
$\cv_{gb}^2$):

\begin{cmathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad (\vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff}) \\
\qquad \wedge~ ((\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff})
\end{array}
\end{cmathpar}

\noindent Intuitively, this prevents both the \rcf{getBalance} anomalies
described in figs.~\ref{fig:negativeBalanceAnomaly} and
\ref{fig:missingUpdateAnomaly} from ever occurring.

Finally, since there are no restrictions on when or how a \rcf{deposit}
operation can execute, its contract is simply $\small \true$.

@@

Notice that the contracts for \rcf{withdraw} and \rcf{getBalance} only express
application-level consistency requirements, and make no reference to the
semantics of the underlying store. To write contracts, a programmer only needs
to reason about the semantics of the application under the \name system model.
The mapping of application-level consistency requirements to appropriate
store-level guarantees is done automatically behind-the-scene. How might one go
about ensuring that an execution adheres to a contract? The challenge is that a
contract provides a declarative (axiomatic) specification of an execution,
while what is required is an operational procedure for \emph{enforcing} its
implicit constraints.
